---
title: "Chi-Square Goodness of Fit"
---

```{r}
#| include: false
#| cache: false
source("../_colors.R")
```



::: {.chapterintro}
In previous chapters, we have developed inferential methods for comparing proportions across two groups. But what if we want to assess whether a single categorical variable with multiple levels follows a particular distribution? For instance, we might want to know if a jury pool is racially representative of the community, or whether the days of the week on which accidents occur are equally likely. The **chi-square goodness of fit test** is designed for exactly this purpose. It compares a set of observed counts to a set of expected counts derived from a hypothesized distribution, using a single test statistic that summarizes all of the differences at once.
:::

## Observed versus expected counts

Consider data from a random sample of 275 jurors in a small county. Jurors identified their racial group, and we would like to determine if these jurors are racially representative of the county's population of registered voters. The observed counts and the population proportions are shown below.

|  | White | Black | Hispanic | Other | Total |
|---|:---:|:---:|:---:|:---:|:---:|
| Observed jurors | 205 | 26 | 25 | 19 | 275 |
| Population proportion | 0.72 | 0.07 | 0.12 | 0.09 | 1.00 |

While the proportions in the jury sample do not precisely match the population proportions, it is unclear whether these data provide convincing evidence that the sample is not representative. If the jurors really were randomly sampled from the registered voters, we might expect small differences due to chance. However, unusually large differences may provide convincing evidence that the juries were not representative.

### Computing expected counts

If the jurors were randomly sampled from the population (i.e., if the null hypothesis is true), we would expect the sample proportions to roughly match the population proportions. The **expected count** for each category is found by multiplying the sample size by the hypothesized proportion for that category.

::: {.workedexample}
Of the 275 people who served on a jury, about how many would we expect to be White if jurors are randomly selected from the population? How many would we expect to be Black?

---

About 72% of the population is White, so we would expect about 72% of the jurors to be White: $0.72 \times 275 = 198$. Similarly, we would expect about 7% of the jurors to be Black, which would correspond to about $0.07 \times 275 = 19.25$ Black jurors.
:::

::: {.guidedpractice}
Twelve percent of the population is Hispanic and 9% represent other races. How many of the 275 jurors would we expect to be Hispanic? How many from other races?

<details>
<summary>Show answer</summary>
We would expect $0.12 \times 275 = 33$ Hispanic jurors and $0.09 \times 275 = 24.75$ jurors from other races.
</details>
:::

The complete table of observed and expected counts is:

|  | White | Black | Hispanic | Other | Total |
|---|:---:|:---:|:---:|:---:|:---:|
| Observed counts | 205 | 26 | 25 | 19 | 275 |
| Expected counts | 198 | 19.25 | 33 | 24.75 | 275 |

::: {.important}
**Computing expected counts for a goodness of fit test.**

If the null hypothesis specifies proportions $p_1, p_2, \ldots, p_k$ for $k$ categories, and the total sample size is $n$, then the expected count for category $i$ is:

$$E_i = n \times p_i$$
:::

### Setting up hypotheses

The sample proportions from each racial group were not a precise match for the population proportions. We need to test whether the differences are strong enough to provide convincing evidence that the jurors are not a random sample. These ideas can be organized into hypotheses:

- $H_0$: The jurors are a random sample from the population. The observed counts reflect natural sampling fluctuation.
- $H_A$: The jurors are not randomly sampled from the population. There is bias in juror selection.

To evaluate these hypotheses, we need to quantify how different the observed counts are from the expected counts. Strong evidence for the alternative hypothesis would come in the form of unusually large deviations from what would be expected based on sampling variation alone.

## The chi-square statistic {#sec-chisq-gof-statistic}

To build a test statistic, we first compute the standardized difference between each observed and expected count. For each category $i$:

$$Z_i = \frac{O_i - E_i}{\sqrt{E_i}}$$

where $O_i$ is the observed count and $E_i$ is the expected count. The denominator $\sqrt{E_i}$ serves as the standard error for the count.

For the jury data:

$$
\begin{aligned}
Z_{\text{White}} &= \frac{205 - 198}{\sqrt{198}} = 0.50 \\
Z_{\text{Black}} &= \frac{26 - 19.25}{\sqrt{19.25}} = 1.54 \\
Z_{\text{Hispanic}} &= \frac{25 - 33}{\sqrt{33}} = -1.39 \\
Z_{\text{Other}} &= \frac{19 - 24.75}{\sqrt{24.75}} = -1.16
\end{aligned}
$$

We would like to use a single test statistic to determine if these four standardized differences are unusually far from zero as a group. We do this by squaring each standardized difference and adding them up:

$$X^2 = Z_1^2 + Z_2^2 + Z_3^2 + Z_4^2 = 0.50^2 + 1.54^2 + (-1.39)^2 + (-1.16)^2 = 5.89$$

Squaring each standardized difference before adding them together does two things:

1. Any standardized difference that is squared will be positive, so negative and positive deviations both contribute to the test statistic.
2. Differences that are already unusual (e.g., a standardized difference of 2.5) become much larger after squaring, which gives them more weight.

::: {.definition}
**Chi-square test statistic.**

The **chi-square test statistic** is the sum of the squared standardized differences between observed and expected counts:

$$X^2 = \sum_{i=1}^{k} \frac{(O_i - E_i)^2}{E_i}$$

where the sum is over all $k$ categories, $O_i$ is the observed count in category $i$, and $E_i$ is the expected count in category $i$ under the null hypothesis.
:::

::: {.original}
## Simulation-based goodness of fit test

Before turning to the mathematical chi-square distribution, it is instructive to see how a simulation-based approach works for goodness of fit tests.

The idea is the same as in other randomization tests: if the null hypothesis is true (the data come from the hypothesized distribution), we can simulate many datasets that would arise under the null hypothesis and compute the chi-square statistic for each. This gives us a null distribution of the chi-square statistic, against which we can compare our observed statistic.

**How the simulation works:**

1. Under $H_0$, each observation has probability $p_i$ of falling into category $i$.
2. For each simulated sample of size $n$, randomly assign each observation to a category according to the null probabilities.
3. Compute the chi-square statistic $X^2$ for the simulated data.
4. Repeat many times (e.g., 1,000 or 10,000 times) to build a null distribution.
5. Compare the observed $X^2$ to the null distribution. The p-value is the proportion of simulated statistics that are as large or larger than the observed statistic.

For the jury example, each simulated dataset would randomly assign 275 jurors to the four racial categories with probabilities 0.72, 0.07, 0.12, and 0.09. The resulting chi-square statistics from the simulated datasets form the null distribution.

```{r}
#| label: fig-chisq-gof-null
#| fig-cap: "Histogram of chi-square statistics from 1,000 simulations under the null hypothesis for the jury example. The observed statistic of 5.89 is marked. Most simulated values fall below 5.89, but a meaningful fraction exceed it. [Open in StatLens]({{< var tool-url >}}/simulate/goodness-of-fit/){.open-in-statlens target='_blank' aria-label='Simulate a goodness-of-fit chi-square null distribution in StatLens'} — enter the jury counts and null proportions to regenerate this null distribution live."
#| fig-alt: "Histogram of simulated chi-square statistics for the jury goodness of fit test, with the observed value of 5.89 marked by a vertical line."

library(ggplot2)

set.seed(42)
null_probs <- c(0.72, 0.07, 0.12, 0.09)
expected <- 275 * null_probs
n_sims <- 1000

chisq_stats <- replicate(n_sims, {
  sim_counts <- as.numeric(table(factor(
    sample(1:4, size = 275, replace = TRUE, prob = null_probs),
    levels = 1:4)))
  sum((sim_counts - expected)^2 / expected)
})

sim_df <- data.frame(chisq = chisq_stats)

ggplot(sim_df, aes(x = chisq)) +
  geom_histogram(binwidth = 0.5, fill = COL_BLUE, color = "white", alpha = 0.8) +
  geom_vline(xintercept = 5.89, color = COL_RED, linewidth = 1, linetype = "dashed") +
  annotate("text", x = 5.89, y = Inf, label = expression(X^2 == 5.89),
           hjust = -0.15, vjust = 2, size = 4, color = COL_RED) +
  labs(x = expression(Simulated ~ X^2 ~ Statistic), y = "Count") +
  theme_minimal(base_size = 14)
```

The simulation-based approach confirms what we will find with the mathematical model: the observed chi-square statistic of 5.89 is not unusual enough to reject the null hypothesis at a 0.05 discernibility level.

::: {.statlens}
**See it in action.** Open the [Chi-Square GOF walkthrough]({{< var tool-url >}}/inference/chisq/?activity=chisq-goodness-of-fit.json){target="_blank"} for a guided pass through framing → checking expected counts → running the test → reading per-category contributions. Or open the [Goodness-of-Fit Simulation]({{< var tool-url >}}/simulate/goodness-of-fit/?activity=goodness-of-fit-simulation.json){target="_blank"} to build the null χ² distribution by repeatedly drawing samples from the hypothesized distribution — a **Predict → Do → Explain** activity that lets you see *why* the mathematical chi-square curve is a reasonable model. For a bare tool, open the [Chi-Square Test tool]({{< var tool-url >}}/inference/chisq/){target="_blank"} (analytic) or the [Goodness-of-Fit Simulator]({{< var tool-url >}}/simulate/goodness-of-fit/){target="_blank"} (simulation) directly on your own data.
:::
:::

## The chi-square distribution {#sec-chisq-distribution}

The chi-square test statistic $X^2$ follows a **chi-square distribution** when the null hypothesis is true and the sample size conditions are met. The chi-square distribution is characterized by a single parameter called the **degrees of freedom** (df), which influences the shape, center, and spread of the distribution.

```{r}
#| label: fig-chisq-distributions
#| fig-cap: "Three chi-square distributions with degrees of freedom 2, 4, and 9. As the degrees of freedom increase, the distribution becomes more symmetric, the center moves to the right, and the variability increases. [↗ Try this live](https://learnlens.org/statlens/conceptual/chi-square-distribution/) — drive the df slider and watch the shape drift right."
#| fig-alt: "Three overlapping chi-square distribution curves with 2, 4, and 9 degrees of freedom, showing how the shape changes with increasing degrees of freedom."

x_vals <- seq(0.01, 25, length.out = 500)

chisq_df <- data.frame(
  x = rep(x_vals, 3),
  density = c(dchisq(x_vals, df = 2), dchisq(x_vals, df = 4), dchisq(x_vals, df = 9)),
  df = factor(rep(c("df = 2", "df = 4", "df = 9"), each = length(x_vals)),
              levels = c("df = 2", "df = 4", "df = 9"))
)

ggplot(chisq_df, aes(x = x, y = density, color = df, linetype = df)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("df = 2" = COL_BLUE, "df = 4" = COL_PINK, "df = 9" = COL_GREEN),
                     name = "Degrees of Freedom") +
  scale_linetype_manual(values = c("df = 2" = "solid", "df = 4" = "dashed", "df = 9" = "dotted"),
                        name = "Degrees of Freedom") +
  labs(x = expression(X^2), y = "Density") +
  coord_cartesian(ylim = c(0, 0.5)) +
  theme_minimal(base_size = 14)
```

Key properties of the chi-square distribution:

- It is always non-negative (the chi-square statistic is a sum of squared values).
- It is right-skewed, especially for small degrees of freedom.
- As the degrees of freedom increase, the distribution becomes more symmetric and shifts to the right.
- The mean of the distribution equals the degrees of freedom.

### Degrees of freedom for goodness of fit

For a goodness of fit test with $k$ categories, the degrees of freedom are:

$$df = k - 1$$

::: {.workedexample}
How many categories were there in the juror example? How many degrees of freedom should be associated with the chi-square distribution?

---

There were $k = 4$ categories: White, Black, Hispanic, and Other. The degrees of freedom are $df = k - 1 = 3$.
:::

## Finding a p-value {#sec-chisq-gof-pvalue}

If the null hypothesis is true, $X^2$ follows a chi-square distribution with $k - 1$ degrees of freedom. Because larger values of $X^2$ correspond to greater deviations from the null hypothesis, the p-value is always found from the **upper tail** of the distribution.

::: {.workedexample}
For the jury data, the test statistic is $X^2 = 5.89$ and the degrees of freedom are $df = 3$. Find the p-value.

---

Using a chi-square distribution with 3 degrees of freedom, the area to the right of 5.89 is approximately 0.117.

```{r}
#| label: fig-chisq-pvalue-gof
#| fig-cap: "Chi-square distribution with 3 degrees of freedom, with the area above 5.89 shaded to represent the p-value of approximately 0.117. [↗ Try this live](https://learnlens.org/statlens/conceptual/chi-square-distribution/) — drag the cutline to any χ² value and read the p-value."
#| fig-alt: "Chi-square distribution with 3 degrees of freedom. The area to the right of 5.89 is shaded, representing the p-value of approximately 0.117."

x_vals <- seq(0.01, 18, length.out = 500)
y_vals <- dchisq(x_vals, df = 3)
curve_df <- data.frame(x = x_vals, y = y_vals)

shade_x <- x_vals[x_vals >= 5.89]
shade_df <- data.frame(
  x = c(5.89, shade_x, max(shade_x)),
  y = c(0, dchisq(shade_x, df = 3), 0)
)

ggplot() +
  geom_polygon(data = shade_df, aes(x = x, y = y), fill = COL_PINK, alpha = 0.5) +
  geom_line(data = curve_df, aes(x = x, y = y), color = COL_BLUE, linewidth = 1.1) +
  geom_vline(xintercept = 5.89, linetype = "dashed", color = COL_RED) +
  annotate("text", x = 8, y = 0.08, label = "p-value = 0.117", size = 4.5, color = COL_RED) +
  labs(x = expression(X^2), y = "Density") +
  theme_minimal(base_size = 14)
```

Since the p-value of 0.117 is larger than $\alpha = 0.05$, we do not reject the null hypothesis. The data do not provide convincing evidence of racial bias in the juror selection process.
:::

::: {.workedexample}
Consider a chi-square distribution with 5 degrees of freedom and a test statistic of 5.1. Find the p-value.

---

Using statistical software or a chi-square table, the area to the right of 5.1 on a chi-square distribution with 5 degrees of freedom is approximately 0.404. This is a very large p-value, so we would not reject the null hypothesis.
:::

::: {.guidedpractice}
A chi-square distribution with 4 degrees of freedom has a test statistic of 10. Find the approximate p-value.

<details>
<summary>Show answer</summary>
Using statistical software, the area to the right of 10 on a chi-square distribution with 4 degrees of freedom is approximately 0.040. Since this is less than 0.05, we would reject the null hypothesis.
</details>
:::

## Chi-square goodness of fit test

We can now summarize the complete procedure for a chi-square goodness of fit test.

::: {.important}
**Chi-square test for goodness of fit.**

Suppose we want to evaluate whether the observed counts $O_1, O_2, \ldots, O_k$ in $k$ categories are consistent with a hypothesized distribution that specifies expected counts $E_1, E_2, \ldots, E_k$.

**Hypotheses:**

- $H_0$: The data follow the hypothesized distribution (the observed counts are consistent with the expected counts, up to sampling variability).
- $H_A$: The data do not follow the hypothesized distribution.

**Test statistic:**

$$X^2 = \sum_{i=1}^{k} \frac{(O_i - E_i)^2}{E_i}$$

When the null hypothesis is true and the conditions are met, $X^2$ follows a chi-square distribution with $df = k - 1$.

The p-value is found from the upper tail of the chi-square distribution.

**Conditions:**

- **Independence.** Each case that contributes a count must be independent of all other cases.
- **Sample size.** Each expected count must be at least 5.
:::

## Evaluating goodness of fit for a distribution

The goodness of fit test can also be used to evaluate whether data follow a particular theoretical distribution.

::: {.workedexample}
Suppose we observe the daily change in a stock index over 1,362 trading days. We classify each day according to how many consecutive days of increase preceded it, and compare the observed frequencies to what we would expect if each day's movement were independent of the previous day's (a geometric distribution model). The observed and expected counts are shown below.

| Days since last increase | 1 | 2 | 3 | 4 | 5 | 6 | 7+ |
|---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Observed | 717 | 369 | 155 | 69 | 28 | 14 | 10 |
| Expected | 743 | 338 | 154 | 70 | 32 | 14 | 12 |

Are these data consistent with the geometric distribution model?

---

We set up the hypotheses:

- $H_0$: Stock returns are independent from one day to the next and follow a geometric model.
- $H_A$: Stock returns show a pattern inconsistent with the geometric model.

The chi-square statistic is:

$$X^2 = \frac{(717 - 743)^2}{743} + \frac{(369 - 338)^2}{338} + \frac{(155 - 154)^2}{154} + \frac{(69 - 70)^2}{70} + \frac{(28 - 32)^2}{32} + \frac{(14 - 14)^2}{14} + \frac{(10 - 12)^2}{12} = 4.61$$

With $k = 7$ categories, $df = 6$. The p-value for $X^2 = 4.61$ with 6 degrees of freedom is approximately 0.595.

Since the p-value is very large, we do not reject the null hypothesis. The data are consistent with the geometric distribution model, suggesting that stock movements may be independent from day to day over this period.
:::

## Chapter review {#sec-chisq-gof-review}

### Summary

In this chapter, we developed the chi-square goodness of fit test, which assesses whether a sample of categorical data is consistent with a hypothesized distribution. The test uses the chi-square statistic, which measures how far the observed counts deviate from the expected counts. When the null hypothesis is true and the conditions are met, this statistic follows a chi-square distribution with $k - 1$ degrees of freedom, where $k$ is the number of categories. The p-value is always computed from the upper tail of the chi-square distribution, since larger values indicate greater discrepancy between the observed and expected counts.

### Key terms

- Expected count
- Chi-square statistic ($X^2$)
- Chi-square distribution
- Degrees of freedom (goodness of fit)
- Goodness of fit test

## Exercises {.exercises}

{{< include ../exercises/_chisq-gof-exercises.qmd >}}

{{< include ../exercises/_chisq-gof-statlens-exercises.qmd >}}
