---
title: "Inference for Proportions"
---

```{r}
#| include: false
#| cache: false
source("../_colors.R")
```



::: {.chapterintro}
In earlier chapters, we used simulation-based methods to draw conclusions about population proportions. In the randomization tests and bootstrap confidence intervals chapters, we built randomization distributions and bootstrap confidence intervals; in the normal approximation chapter, we learned that the normal distribution provides a mathematical model for these sampling distributions when certain conditions are met. In this chapter, we bring everything together for proportions. We develop normal-based confidence intervals and hypothesis tests for a single proportion and for the difference between two proportions. Throughout, we compare these formula-based results to the simulation-based results from earlier chapters, reinforcing the idea that these are two routes to the same destination.
:::

## Case study: is the complication rate lower? {#sec-medical-consultant-setup}

People providing an organ for donation sometimes seek the help of a special **medical consultant**. These consultants assist the patient in all aspects of the surgery, with the goal of reducing the possibility of complications during the medical procedure and recovery.

One consultant tried to attract patients by noting the national complication rate for liver donor surgeries is about 10%, but her clients have had only 3 complications in the 62 liver donor surgeries she has facilitated. She claims this is strong evidence that her work reduces complications.

Let $p$ denote the true complication rate for this consultant's clients. The point estimate is
$$\hat{p} = \frac{3}{62} = 0.048.$$

The sample rate 4.8% is well below the national 10%. But we saw the same puzzle in the sampling variability chapter: even when nothing is going on, sample proportions vary from sample to sample. **Could a rate as low as 4.8% out of 62 have happened by chance, if the true rate really were 10%?** That is the question a hypothesis test is built to answer:

- $H_0$: The consultant's true complication rate matches the national rate, $p = 0.10$.
- $H_A$: The consultant's true complication rate is lower, $p < 0.10$.

To weigh the evidence, we need to know how much $\hat{p}$ would vary from sample to sample **if $H_0$ were true**. That variability is captured by the null distribution of $\hat{p}$.

## Building the null distribution by simulation {#sec-null-dist-buildup}

We do not need calculus or a formula to picture the null distribution --- we can build it directly with simulation. This mirrors the logic of the randomization tests chapter, applied here to a single proportion.

### The marbles metaphor

Imagine a large bag of marbles: 10% orange (complication) and 90% blue (no complication). This bag models the world under $H_0$. Each surgery is one draw from the bag; a "study" of $n = 62$ surgeries is one scoop of 62 marbles. The proportion of orange marbles in the scoop is one simulated $\hat{p}_{sim}$ from the null distribution.

If we scoop 62 marbles and get 5 orange and 57 blue, then
$$\hat{p}_{sim1} = \frac{5}{62} = 0.081.$$

That is *one* simulated study. Under $H_0$, this could easily have happened, even though it does not equal 0.10 exactly.

::: {.workedexample}
Is one simulated $\hat{p}_{sim}$ enough to draw a conclusion about the consultant?

---

No. One draw tells us nothing about the *shape* of the null distribution --- it is a single value that happens to land somewhere by chance. To know what values of $\hat{p}$ are typical and which are extreme under $H_0$, we need many simulations.
:::

### From one draw to many

The natural next question is: **what happens as we accumulate more draws?** Does the null distribution take a recognizable shape? Does it settle down or keep drifting?

This is the moment where a static picture struggles. Jumping from "one draw at $\hat{p}_{sim1} = 0.081$" to a smooth histogram of 10,000 draws is a big cognitive leap when you have not seen this before. It is worth pausing to build the distribution one increment at a time.

::: {.statlens}
**Predict → Do → Explain.** Open the [Sampling Distribution Lab (proportion mode)]({{< var tool-url >}}/conceptual/sampling-lab/?activity=sampling-distribution-proportion.json&p=0.10&n=62){target="_blank"} --- preloaded with $p = 0.10$ and $n = 62$, the null-hypothesis world for the medical-consultant case. The gated activity opens with a **Predict** step (commit to where the distribution of $\hat{p}_{sim}$ will be centered, how spread out, and what shape), walks you through drawing 1, 10, 100, and 1,000 samples in sequence, and closes with an **Explain** step where you articulate what the buildup showed. Your responses save locally in the tool, so you can return to your work on the same device.
:::

Three ideas the activity lands on --- worth flagging in the reading so you can compare them to your own conclusions:

- More samples do **not** make the individual $\hat{p}_{sim}$ values more precise --- they fill in the *same* null distribution. To get narrower $\hat{p}_{sim}$ values you must increase $n$ (the size of each scoop), not the number of scoops.
- The null distribution has a stable center at $p = 0.10$ and a stable spread even while the individual draws stay noisy. That stability is what makes the p-value calculation possible.
- The distribution is **roughly bell-shaped, but with a mild right skew**. That skew is not noise --- it is a signal. With $p = 0.10$ and $n = 62$, the expected number of successes is $np = 6.2 < 10$, so the success--failure condition fails and the normal approximation is only approximate. When $p$ sits near 0 (or 1) and $n$ is not large, $\hat{p}$ has more room to wander toward the middle than toward the floor, which pushes a tail out to the right. This is the case where a simulated null distribution is more trustworthy than a normal formula --- exactly the situation we started the chapter with.

### The null distribution at 1,000 simulations

A computer can repeat the marbles scoop as many times as we want. Here are the results of 1,000 simulated studies under $H_0$: each study draws 62 marbles from the 10% / 90% bag, and we record $\hat{p}_{sim}$ from each.

```{r}
#| label: fig-null-dist-medical
#| fig-cap: "The null distribution of p-hat from 1,000 simulated studies under H₀: p = 0.10, n = 62. Simulations with p-hat ≤ 0.048 (as extreme as the consultant's observed value, or more extreme) are shaded orange. [↗ Try this live](https://learnlens.org/statlens/conceptual/sampling-lab/) — simulate under $H_0$ with your own $p$ and $n$ and rebuild the null."
#| fig-alt: "Histogram of 1,000 simulated sample proportions under a null proportion of 0.10 with sample size 62. The distribution is centered near 0.10 and roughly bell-shaped but slightly right-skewed. The left tail at or below 0.048 is shaded."
#| fig-width: 7
#| fig-height: 3.5
#| echo: false
library(ggplot2)
set.seed(334422)
null_df <- data.frame(phat = rbinom(1000, size = 62, prob = 0.10) / 62)
tail_share <- mean(null_df$phat <= 0.0484)

ggplot(null_df, aes(x = phat, fill = phat <= 0.0484)) +
  geom_histogram(binwidth = 1/62, color = "white", show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = COL_BLUE, "TRUE" = COL_ORANGE)) +
  geom_vline(xintercept = 0.0484, linetype = "dashed", color = COL_ORANGE, linewidth = 0.7) +
  annotate("text", x = 0.048, y = Inf,
           label = expression(hat(p) == 0.048),
           vjust = 1.4, hjust = 1.1, color = COL_ORANGE, size = 3.6, fontface = "bold") +
  labs(x = expression(hat(p)[sim]),
       y = "Count (out of 1,000)") +
  theme_minimal(base_size = 13)
```

Two observations to record before we compute anything:

1. **The distribution is centered near $p = 0.10$.** That is not luck --- it is because we built it under $H_0$, where the truth is $p = 0.10$. Simulated $\hat{p}$ values scatter symmetrically-ish around that center.
2. **The distribution has a shape we can measure.** It looks roughly bell-shaped but with a mild right skew, because $p = 0.10$ is close to 0 and $\hat{p}$ cannot go below zero.

### Reading the p-value

The consultant's observed $\hat{p} = 0.048$ is marked with the dashed line. The orange-shaded bars are the simulated $\hat{p}_{sim}$ values that are *at least as extreme* as the observed value in the direction of the alternative --- here, $\hat{p}_{sim} \leq 0.048$.

The p-value is the fraction of the simulated studies that landed in the shaded tail:

$$\text{p-value} \;=\; \frac{\text{simulations with } \hat{p}_{sim} \leq 0.048}{1{,}000} \;\approx\; `r round(tail_share, 3)`.$$

::: {.workedexample}
Using $\alpha = 0.05$, what conclusion should we draw?

---

The p-value $\approx `r round(tail_share, 3)`$ is greater than $\alpha = 0.05$. We **fail to reject** $H_0$. The data do not provide convincing evidence that the consultant's true complication rate is lower than the national 10% --- a rate of 3 in 62 could plausibly occur even if her true rate were 10%.

Failing to reject $H_0$ is not the same as concluding that her rate *equals* 10%. It only says the data are compatible with the national rate; a larger study would be needed to distinguish a small real improvement from chance.
:::

::: {.important}
**Simulating the null distribution of $\hat{p}$.**

Whenever we want to test $H_0: p = p_0$ but a normal-based formula is unreliable, we can build the null distribution directly:

1. Set up a marbles-bag with the proportion $p_0$ specified by $H_0$.
2. Repeatedly scoop $n$ marbles and record the simulated $\hat{p}_{sim}$.
3. Read the p-value as the tail area of the null distribution beyond the observed $\hat{p}$.

This works for **any** sample size and does not require the success--failure condition. The formula-based approach in the next section is faster when its conditions are met, but the simulation is always available as a backup.
:::

## Normal approximation for proportions {#sec-normal-approx-prop}

### The sampling distribution of $\hat{p}$

In the normal approximation chapter, we saw that the Central Limit Theorem guarantees that the sampling distribution of $\hat{p}$ is approximately normal when certain conditions are met. Let's state these conditions precisely.

::: {.important}
**Sampling distribution of $\hat{p}$.**

The sampling distribution of the sample proportion $\hat{p}$, based on a sample of size $n$ from a population with true proportion $p$, is approximately normal when:

1. **Independence.** The observations are independent (e.g., from a simple random sample).
2. **Success-failure condition.** We expect to see at least 10 successes and 10 failures: $np \geq 10$ and $n(1-p) \geq 10$.

When these conditions are met, the sampling distribution is approximately:

$$\hat{p} \sim N\!\left(p,\ \sqrt{\frac{p(1-p)}{n}}\right)$$
:::

::: {.definition}
The **success-failure condition** requires that the expected number of successes ($np$) and the expected number of failures ($n(1-p)$) are both at least 10. This ensures the sampling distribution of $\hat{p}$ is not too skewed for the normal approximation to work well.
:::

::: {.workedexample}
A medical consultant facilitated 62 liver transplant surgeries, with 3 complications. Under the null hypothesis that the true complication rate is $p_0 = 0.10$, is it appropriate to use the normal approximation?

---

Check the success-failure condition using $p_0 = 0.10$:

- Expected successes (complications): $np_0 = 62 \times 0.10 = 6.2$
- Expected failures: $n(1-p_0) = 62 \times 0.90 = 55.8$

The expected number of successes (6.2) is less than 10. The **success-failure condition is not met**, so the normal approximation should not be used here. A simulation-based approach (such as the parametric bootstrap from the bootstrap confidence intervals chapter) would be more appropriate.
:::

::: {.guidedpractice}
A poll surveys 826 randomly selected payday loan borrowers and finds that 51% support a new regulation. Check whether the normal approximation can be used to model $\hat{p}$ for a hypothesis test with $H_0: p = 0.50$.

<details>
<summary>Show answer</summary>
Check the success-failure condition using $p_0 = 0.50$: $np_0 = 826 \times 0.50 = 413$ and $n(1-p_0) = 826 \times 0.50 = 413$. Both are well above 10, so the success-failure condition is met. Combined with the random sample (independence), the normal approximation is appropriate.
</details>
:::

### The standard error of $\hat{p}$

::: {.important}
**Standard error of $\hat{p}$.**

When conditions are met, the variability of $\hat{p}$ is described by:

$$SE(\hat{p}) = \sqrt{\frac{p(1-p)}{n}}$$

Since we rarely know the true $p$:

- **For confidence intervals**, use $\hat{p}$ as the best guess: $SE = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$
- **For hypothesis tests**, use $p_0$ (the value from $H_0$): $SE = \sqrt{\frac{p_0(1-p_0)}{n}}$
:::

::: {.guidedpractice}
Consider polls of size 300 where approximately 2/3 of voters support legalized marijuana. Calculate the standard error of $\hat{p}$.

<details>
<summary>Show answer</summary>
Using $p \approx 2/3$: $SE = \sqrt{\frac{(2/3)(1/3)}{300}} = \sqrt{\frac{2/9}{300}} = \sqrt{0.000741} = 0.027$.
</details>
:::

## Confidence interval for a single proportion {#sec-ci-one-prop}

### The formula

::: {.important}
**Confidence interval for a single proportion $p$.**

$$\hat{p} \pm z^* \times \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

where $z^*$ is the critical value for the desired confidence level (e.g., $z^* = 1.96$ for 95%).

**Steps:**

1. Check conditions: independence and success-failure using $\hat{p}$.
2. Compute $SE$ using $\hat{p}$.
3. Apply the confidence interval formula.
:::

### Worked example: Payday loan regulation

::: {.workedexample}
A simple random sample of 826 payday loan borrowers was surveyed, and 70% supported new regulations on payday lenders. Construct a 95% confidence interval for the true proportion who support the regulations.

---

**Step 1: Check conditions.**

- *Independence.* The data are a random sample.
- *Success-failure condition.* Using $\hat{p} = 0.70$: $n\hat{p} = 826 \times 0.70 = 578$ and $n(1-\hat{p}) = 826 \times 0.30 = 248$. Both are well above 10.

**Step 2: Compute SE.**

$$SE = \sqrt{\frac{0.70 \times 0.30}{826}} = \sqrt{\frac{0.21}{826}} = \sqrt{0.000254} = 0.016$$

**Step 3: Construct the interval.**

$$0.70 \pm 1.96 \times 0.016 = 0.70 \pm 0.031 = (0.669, 0.731)$$

We are 95% confident that between 66.9% and 73.1% of all payday loan borrowers support the new regulation.
:::

### Changing the confidence level

The confidence level is controlled by the choice of $z^*$:

| Confidence level | $z^*$ | Interval width |
|-----------------|-------|---------------|
| 90% | 1.645 | Narrower |
| 95% | 1.960 | Moderate |
| 99% | 2.576 | Wider |

: Critical values for common confidence levels. {.striped}

Higher confidence requires a wider interval -- we must cast a wider net to be more confident we have captured the true parameter.

::: {.guidedpractice}
Using the payday loan data ($\hat{p} = 0.70$, $n = 826$, $SE = 0.016$), construct a 99% confidence interval.

<details>
<summary>Show answer</summary>
Using $z^* = 2.576$: $0.70 \pm 2.576 \times 0.016 = 0.70 \pm 0.041 = (0.659, 0.741)$. The 99% interval is wider than the 95% interval, as expected.
</details>
:::

::: {.guidedpractice}
A study on stent surgery found a point estimate of 0.090 (the difference in stroke rates) with $SE = 0.028$. Construct a 90% confidence interval.

<details>
<summary>Show answer</summary>
Using $z^* = 1.645$: $0.090 \pm 1.645 \times 0.028 = 0.090 \pm 0.046 = (0.044, 0.136)$. We are 90% confident that stent implantation increases the 30-day stroke rate by between 4.4% and 13.6%.
</details>
:::

::: {.important}
**General confidence interval formula.** If a point estimate follows a normal model with standard error $SE$, then a confidence interval for the population parameter is:

$$\text{point estimate} \pm z^* \times SE$$

where $z^*$ corresponds to the chosen confidence level.
:::

::: {.callout-tip}
### Planning a study: choosing $n$

Before collecting data, researchers often need to decide how large a sample to draw. The width of the CI above is driven by the standard error $\sqrt{\hat{p}(1-\hat{p})/n}$, so hitting a target margin of error is really a question of choosing $n$ large enough. We work through the arithmetic (with worst-case $\hat{p} = 0.5$ and other refinements) in the [sample size chapter](sample-size.qmd) later in the book.
:::

## Hypothesis test for a single proportion {#sec-ht-one-prop}

### The Z test for a proportion

::: {.important}
**The test statistic for a single proportion is a Z.**

$$Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0) / n}}$$

where $\hat{p}$ is the sample proportion and $p_0$ is the hypothesized proportion from $H_0$.

When $H_0$ is true and conditions are met, $Z$ follows a standard normal distribution $N(0, 1)$.

**Conditions:**

- Independent observations
- $np_0 \geq 10$ and $n(1 - p_0) \geq 10$
:::

Note that for hypothesis tests, the standard error is computed using $p_0$ (not $\hat{p}$), because we are evaluating the data under the assumption that $H_0$ is true.

### Worked example: Support for credit checks

::: {.workedexample}
From a random sample of 826 payday loan borrowers, 51% said they would support a regulation requiring lenders to pull credit reports. Is there convincing evidence that a majority (more than 50%) of borrowers support this regulation?

---

**Step 1: State hypotheses.**

- $H_0$: $p = 0.50$ (no majority support)
- $H_A$: $p > 0.50$ (majority support)

This is a one-sided test.

**Step 2: Check conditions.**

- *Independence.* Random sample.
- *Success-failure.* Using $p_0 = 0.50$: $np_0 = 826 \times 0.50 = 413$ and $n(1-p_0) = 413$. Both $\geq 10$.

**Step 3: Compute the test statistic and p-value.**

$$SE = \sqrt{\frac{0.50 \times 0.50}{826}} = \sqrt{\frac{0.25}{826}} = 0.017$$

$$Z = \frac{0.51 - 0.50}{0.017} = \frac{0.01}{0.017} = 0.59$$

```{r}
#| label: fig-normal-right-tail-z059
#| fig-cap: "A standard normal curve with the area to the right of Z = 0.59 shaded, representing the one-sided p-value. [↗ Try this live](https://learnlens.org/statlens/conceptual/normal-distribution/) — drag the cutline to any Z value and read off the right-tail area."
#| fig-alt: "A standard normal distribution curve with the area to the right of Z equals 0.59 shaded. This shaded area represents the p-value for the one-sided test."

library(ggplot2)

x <- seq(-4, 4, length.out = 500)
y <- dnorm(x)
base_df <- data.frame(x = x, y = y)
shade_df <- data.frame(x = x[x >= 0.59], y = dnorm(x[x >= 0.59]))

ggplot(base_df, aes(x = x, y = y)) +
  geom_area(data = shade_df, aes(x = x, y = y), fill = COL_PINK, alpha = 0.5) +
  geom_line(linewidth = 1, color = COL_GREEN) +
  geom_vline(xintercept = 0.59, linetype = "dashed", color = COL_RED) +
  annotate("text", x = 0.59, y = -0.015, label = "Z = 0.59", size = 4, color = COL_RED) +
  annotate("text", x = 2.0, y = 0.08, label = "p-value = 0.278", size = 4, fontface = "bold", color = COL_PINK) +
  labs(x = "Z", y = NULL) +
  theme_minimal(base_size = 14)
```

The area to the right of $Z = 0.59$ is 0.278.

**Step 4: Conclusion.** The p-value of 0.278 is much larger than $\alpha = 0.05$, so we fail to reject $H_0$. The poll does not provide convincing evidence that a majority of payday loan borrowers support the regulation requiring credit checks.
:::

::: {.guidedpractice}
A candidate claims that more than 60% of voters in a district support her. A random sample of 500 voters finds $\hat{p} = 0.64$. Test the claim at $\alpha = 0.05$.

<details>
<summary>Show answer</summary>
Hypotheses: $H_0: p = 0.60$, $H_A: p > 0.60$. Success-failure: $500(0.60) = 300$ and $500(0.40) = 200$, both $\geq 10$. $SE = \sqrt{0.60 \times 0.40 / 500} = \sqrt{0.000480} = 0.0219$. $Z = (0.64 - 0.60)/0.0219 = 1.83$. The right-tail area is 0.034. Since $0.034 < 0.05$, we reject $H_0$. There is sufficient evidence that more than 60% of voters support the candidate.
</details>
:::

### When conditions fail

::: {.caution}
When the success-failure condition is not met, the normal model can underestimate the variability of $\hat{p}$, leading to inaccurate p-values and confidence intervals. In these cases, use simulation-based methods:

- **Bootstrap hypothesis test**: Simulate samples under $H_0$ by drawing from a population with proportion $p_0$ (the parametric bootstrap from the bootstrap confidence intervals chapter).
- **Bootstrap confidence interval**: Resample from the original data to estimate the variability of $\hat{p}$.
:::

## Difference of two proportions {#sec-diff-two-props}

### The parameter and point estimate

When comparing two groups, the parameter of interest is the difference in population proportions: $p_1 - p_2$. The point estimate is $\hat{p}_1 - \hat{p}_2$.

### Confidence interval for a difference in proportions

::: {.important}
**Confidence interval for $p_1 - p_2$.**

$$(\hat{p}_1 - \hat{p}_2) \pm z^* \times \sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$

**Conditions:**

1. **Independence (extended).** Observations are independent within and between groups.
2. **Success-failure condition (each group separately).** $n_1\hat{p}_1 \geq 10$, $n_1(1-\hat{p}_1) \geq 10$, $n_2\hat{p}_2 \geq 10$, and $n_2(1-\hat{p}_2) \geq 10$.
:::

::: {.workedexample}
In a CPR study, patients who had a heart attack were randomly assigned to receive a blood thinner (treatment) or not (control). The survival rates after 24 hours:

| Group | $n$ | Survived | $\hat{p}$ |
|-------|-----|----------|-----------|
| Treatment | 50 | 14 | 0.280 |
| Control | 40 | 11 | 0.275 |

: CPR study results. {.striped}

Construct a 90% confidence interval for the difference in survival rates, $p_T - p_C$.

---

**Check conditions.**

- *Independence.* Patients were randomly assigned, so groups are independent.
- *Success-failure.* Treatment: $50(0.28) = 14 \geq 10$ and $50(0.72) = 36 \geq 10$. Control: $40(0.275) = 11 \geq 10$ and $40(0.725) = 29 \geq 10$. All conditions met.

**Compute the interval.**

$$\hat{p}_T - \hat{p}_C = 0.280 - 0.275 = 0.005$$

$$SE = \sqrt{\frac{0.280(0.720)}{50} + \frac{0.275(0.725)}{40}} = \sqrt{\frac{0.2016}{50} + \frac{0.1994}{40}} = \sqrt{0.004032 + 0.004985} = \sqrt{0.009017} = 0.095$$

Using $z^* = 1.645$ for 90% confidence:

$$0.005 \pm 1.645 \times 0.095 = 0.005 \pm 0.156 = (-0.151, 0.161)$$

We are 90% confident that the difference in 24-hour survival rates is between $-15.1\%$ and $+16.1\%$. Because the interval contains 0, we cannot conclude that the blood thinner affects survival rates.
:::

### Hypothesis test for a difference in proportions

::: {.important}
**The test statistic for comparing two proportions is a Z.**

$$Z = \frac{(\hat{p}_1 - \hat{p}_2) - 0}{\sqrt{\hat{p}_{pool}(1 - \hat{p}_{pool})\left(\frac{1}{n_1} + \frac{1}{n_2}\right)}}$$

where the **pooled proportion** is:

$$\hat{p}_{pool} = \frac{\text{total successes in both groups}}{\text{total sample size}} = \frac{x_1 + x_2}{n_1 + n_2}$$

When $H_0$ is true ($p_1 = p_2$), both groups share the same population proportion, so we pool the data to get a single best estimate.

**Conditions:**

- Independence within and between groups.
- $n_1\hat{p}_{pool} \geq 10$, $n_1(1-\hat{p}_{pool}) \geq 10$, $n_2\hat{p}_{pool} \geq 10$, $n_2(1-\hat{p}_{pool}) \geq 10$.
:::

::: {.workedexample}
Does the blood thinner improve 24-hour survival after CPR? Using the data from the previous example (treatment: 14/50 survived; control: 11/40 survived), test at $\alpha = 0.05$.

---

**Step 1: State hypotheses.**

- $H_0$: $p_T - p_C = 0$ (the blood thinner has no effect)
- $H_A$: $p_T - p_C \neq 0$ (the blood thinner has some effect)

**Step 2: Check conditions.**

The pooled proportion is $\hat{p}_{pool} = \frac{14 + 11}{50 + 40} = \frac{25}{90} = 0.278$. Check:

- $50(0.278) = 13.9 \geq 10$, $50(0.722) = 36.1 \geq 10$
- $40(0.278) = 11.1 \geq 10$, $40(0.722) = 28.9 \geq 10$

Conditions are met.

**Step 3: Compute the test statistic and p-value.**

$$SE = \sqrt{0.278(0.722)\left(\frac{1}{50} + \frac{1}{40}\right)} = \sqrt{0.2007 \times 0.045} = \sqrt{0.009032} = 0.095$$

$$Z = \frac{0.005 - 0}{0.095} = 0.05$$

The two-sided p-value for $Z = 0.05$ is $2 \times P(Z > 0.05) = 2 \times 0.480 = 0.960$.

**Step 4: Conclusion.** The p-value of 0.960 is far larger than 0.05. We fail to reject $H_0$. There is no evidence that the blood thinner affects 24-hour survival rates.
:::

### Why pool for hypothesis tests but not for confidence intervals?

::: {.original}
You may have noticed a subtle difference:

- **Hypothesis tests** use the **pooled proportion** $\hat{p}_{pool}$ in the standard error. This is because under $H_0$, we assume $p_1 = p_2$, so the best estimate of this common proportion comes from combining both samples.
- **Confidence intervals** use the **individual sample proportions** $\hat{p}_1$ and $\hat{p}_2$ separately. We are not assuming the proportions are equal -- we are estimating how different they are.

This distinction applies only to proportions. When we turn to means in the next chapters, we will instead use separate sample standard deviations in both tests and confidence intervals.
:::

## Comparing simulation-based and normal-based methods {#sec-sim-vs-normal-prop}

::: {.original}
Throughout this course, we have developed two parallel approaches to inference:

| Feature | Simulation-based | Normal approximation |
|---------|-----------------|---------------------|
| **Requires conditions?** | Minimal (mainly independence) | Independence + success-failure |
| **Small samples?** | Works well | May be inaccurate |
| **Computational cost** | Higher (needs many simulations) | Lower (one formula) |
| **Conceptual clarity** | High (directly simulates the process) | Moderate (requires distribution theory) |
| **Precision** | Approximate (varies by number of simulations) | Approximate (varies by how well conditions hold) |

: Comparison of simulation-based and normal approximation approaches. {.striped}

### An example comparison

Consider the opportunity cost study from the randomization tests chapter. The observed difference in proportions was 0.20 ($\hat{p}_T - \hat{p}_C$), and we wanted to test $H_0: p_T - p_C = 0$.

**Randomization test (the randomization tests chapter):** We shuffled the group labels 10,000 times. The p-value was approximately 0.006.

**Normal approximation (this chapter):** Using $SE = 0.078$ and $Z = 0.20 / 0.078 = 2.56$, the p-value was 0.005.

Both methods give essentially the same answer and the same conclusion. This is reassuring and typical when conditions are met.

### When do they disagree?

The methods can give different answers when:

1. **The success-failure condition fails.** The normal approximation may undercount the probability in the tails, leading to p-values that are too small. The simulation approach remains valid.
2. **The sampling distribution is noticeably skewed.** The normal model is symmetric by definition, while the true sampling distribution of $\hat{p}$ near 0 or 1 is skewed.

The medical consultant example illustrates this: with only 6.2 expected complications, the normal model gave a p-value of 0.085, while the simulation gave 0.122 -- nearly 45% higher. The simulation-based p-value is more trustworthy in this case.

### Practical recommendation

- **When conditions are met**: Both methods work. Use the normal approximation for its simplicity, or use simulation for its conceptual directness.
- **When conditions are borderline or not met**: Use simulation. The bootstrap and randomization approaches make fewer assumptions and are more reliable in these settings.
- **Always check conditions first.** The success-failure condition is a simple, quick check that helps you decide which approach to use.
:::

## Explore with StatLens {#sec-inference-proportions-statlens}

The [Sampling Distribution Lab (proportion mode)]({{< var tool-url >}}/conceptual/sampling-lab/?activity=sampling-distribution-proportion.json){target="_blank"} builds the sampling distribution of $\hat{p}$ one sample at a time. Set $p = 0.60$ and $n = 50$, watch the distribution fill in, then drop $p$ to $0.05$ to see what happens when the success--failure condition fails.

Analytical tools:

- [One-Proportion $z$ test + CI]({{< var tool-url >}}/inference/one-prop/){target="_blank"}, [Two-Proportion $z$ test + CI]({{< var tool-url >}}/inference/two-props/){target="_blank"}.

Simulation counterparts (for when conditions fail):

- [Bootstrap CI for One Proportion]({{< var tool-url >}}/simulate/bootstrap-prop/){target="_blank"}, [Bootstrap CI for Difference in Proportions]({{< var tool-url >}}/simulate/bootstrap-two-props/){target="_blank"}.
- [Randomization test for One Proportion]({{< var tool-url >}}/simulate/randomization-one-prop/){target="_blank"}, [Randomization test for Two Proportions]({{< var tool-url >}}/simulate/randomization-diff-props/){target="_blank"}.

## Chapter review {#sec-inference-proportions-review}

### Summary

This chapter developed normal-based inference methods for proportions:

- **Success-failure condition.** The normal approximation for $\hat{p}$ requires at least 10 expected successes and 10 expected failures. For confidence intervals, check with $\hat{p}$; for hypothesis tests, check with $p_0$.
- **CI for one proportion**: $\hat{p} \pm z^* \times \sqrt{\hat{p}(1-\hat{p})/n}$. Use $\hat{p}$ in the SE.
- **Hypothesis test for one proportion**: $Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$. Use $p_0$ in the SE.
- **CI for difference of proportions**: $(\hat{p}_1 - \hat{p}_2) \pm z^* \times SE$, where $SE$ uses separate sample proportions.
- **Hypothesis test for difference of proportions**: $Z = \frac{\hat{p}_1 - \hat{p}_2}{SE}$, where $SE$ uses the **pooled proportion** $\hat{p}_{pool}$.
- **Confidence level** is controlled by $z^*$: higher confidence means wider intervals.
- **When conditions fail** (especially the success-failure condition), use simulation-based methods from the randomization tests and bootstrap confidence intervals chapters instead.
- **Normal-based and simulation-based methods agree** when conditions are met and can give different answers when conditions are violated.

### Key terms

Success-failure condition, standard error of $\hat{p}$, Z test for a proportion, pooled proportion, confidence interval for a proportion, confidence interval for a difference of proportions, margin of error, critical value ($z^*$), confidence level.

## Exercises {.exercises}

Answers to odd-numbered exercises are provided in the [Exercise Solutions appendix](../appendices/exercise-solutions.qmd#sec-solutions-inference-proportions) at the back of the book.

{{< include ../exercises/_inference-proportions-exercises.qmd >}}

{{< include ../exercises/_inference-proportions-statlens-exercises.qmd >}}
