---
title: "More About the Normal Distribution"
---

```{r}
#| include: false
#| cache: false
source("../_colors.R")
```


::: {.callout-note appearance="simple" icon=false}
**Builds on the normal-approximation chapter.** This chapter assumes you've already met the normal distribution as it was developed in the [normal-approximation chapter](normal-approximation.qmd) --- the $N(\mu, \sigma)$ notation, the 68-95-99.7 rule, z-scores, and finding tail probabilities and percentiles. Here we look at two extensions that didn't fit earlier: **how to check whether data actually are approximately normal**, and **which real-world variables the normal model fits well (and which it doesn't).**
:::

::: {.chapterintro}
The normal distribution is powerful precisely because so many real-world variables are approximately normal. But *approximately* is a word doing a lot of work --- some variables really are close to normal, others are clearly not, and it matters which is which before you use the normal model to compute probabilities or run a formula-based test. This chapter gives you two tools for making that call.
:::


## Checking whether data are normal {#sec-checking-normality}

Before using the normal distribution as a model, we should check whether the assumption of normality is reasonable. Two visual methods are commonly used, and they complement each other.

### Method 1: Histogram with a normal curve overlay

The simplest approach is to plot a histogram of the data and overlay the best-fitting normal curve (using $\bar{x}$ for $\mu$ and $s$ for $\sigma$). If the histogram closely matches the curve, the normal model is reasonable. If the histogram is skewed, bimodal, or has long tails the curve doesn't capture, the model is a poor fit.

This works well when $n$ is large enough for the histogram to reveal the true shape --- typically $n \geq 50$ or so. With smaller samples, the histogram is noisy and even genuinely normal data can look ragged.

### Method 2: Normal probability plots (QQ plots)

::: {.definition}
A **normal probability plot** (also called a **QQ plot**, short for "quantile-quantile plot") plots the observed data values against the values we would expect if the data were perfectly normal. If the data are approximately normal, the points will fall close to a straight line.
:::

**How to read a QQ plot:**

- **Points on or near the line**: The data are consistent with a normal distribution.
- **Points curving up at the right end**: The distribution is **right-skewed** (a long upper tail).
- **Points curving down at the left end**: The distribution is **left-skewed** (a long lower tail).
- **Points curving away from the line at both ends**: The distribution has **heavier tails** than normal --- more extreme values than the normal predicts.

@fig-qq-plots shows the four patterns you will see most often.

```{r}
#| label: fig-qq-plots
#| fig-cap: "Four QQ plots illustrating common patterns: (a) approximately normal data — points fall close to the line; (b) right-skewed data — points curve upward at the right; (c) left-skewed data — points curve downward at the left; (d) heavy-tailed data — points curve away from the line at both ends."
#| fig-alt: "A two-by-two grid of QQ plots. Top-left shows points following the diagonal line closely. Top-right shows points curving above the line at the right end. Bottom-left shows points curving below the line at the left end. Bottom-right shows points curving away from the line at both the lower-left and upper-right ends."
#| fig-width: 6.5
#| fig-height: 6.5

set.seed(42)
par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1), pty = "s")

# (a) Normal data
normal_data <- rnorm(400)
qqnorm(normal_data, main = "(a) Normal", pch = 16, cex = 0.5,
       col = COL_BLUE, cex.main = 1.3, cex.lab = 1.1)
qqline(normal_data, col = COL_RED, lwd = 1.2, lty = 2)

# (b) Right-skewed data
skewed_data <- rexp(400, rate = 1)
qqnorm(skewed_data, main = "(b) Right-skewed", pch = 16, cex = 0.5,
       col = COL_BLUE, cex.main = 1.3, cex.lab = 1.1)
qqline(skewed_data, col = COL_RED, lwd = 1.2, lty = 2)

# (c) Left-skewed data
left_skewed <- -rexp(400, rate = 1)
qqnorm(left_skewed, main = "(c) Left-skewed", pch = 16, cex = 0.5,
       col = COL_BLUE, cex.main = 1.3, cex.lab = 1.1)
qqline(left_skewed, col = COL_RED, lwd = 1.2, lty = 2)

# (d) Heavy-tailed data
heavy_data <- rt(400, df = 3)
qqnorm(heavy_data, main = "(d) Heavy-tailed", pch = 16, cex = 0.5,
       col = COL_BLUE, cex.main = 1.3, cex.lab = 1.1)
qqline(heavy_data, col = COL_RED, lwd = 1.2, lty = 2)
```

::: {.important}
**Sample size matters.** With small sample sizes ($n < 30$), we should expect more variability in the QQ plot even when the data truly come from a normal distribution. Deviations must be more severe to be considered evidence against normality when $n$ is small.
:::

::: {.guidedpractice}
A researcher collects data and creates a QQ plot. The points follow the diagonal line closely in the middle of the distribution but curve upward at the right end. What does this suggest about the shape of the data?

<details>
<summary>Show answer</summary>
The QQ plot suggests the data are right-skewed. The bulk of the data are approximately normal, but there are larger values in the upper tail than the normal distribution would predict.
</details>
:::


## When to use --- and not use --- the normal model {#sec-when-normal}

The normal distribution is a good model for data that are:

- **Symmetric**: The histogram is roughly mirror-image around the center.
- **Unimodal**: There is a single peak.
- **Bell-shaped**: The histogram has the characteristic tapered shape of the normal curve.
- **No extreme outliers**: Outliers distort both the mean and standard deviation, making the normal model a poor fit.

Many variables are approximately normal:

- Standardized test scores (SAT, ACT, GRE)
- Heights of adults (within a sex)
- Measurement errors in scientific instruments
- Many biological measurements (blood pressure, cholesterol levels)
- Averages of many small independent effects --- which is why sample means behave normally by the Central Limit Theorem

Many variables are **not** normal:

- **Income distributions** --- strongly right-skewed; a few very high earners pull the tail out.
- **Waiting times** --- typically right-skewed, often bounded below by zero.
- **Count data** --- usually better modeled by the binomial or Poisson, not the normal.
- **Proportions and percentages** --- bounded between 0 and 1, so they can't be strictly normal.
- **Ages** --- bounded below by 0 and often bounded above by natural limits, so the tails don't behave like a normal.

::: {.important}
**When the normal model doesn't fit, don't force it.** For skewed data, consider a transformation (e.g., log for right-skewed positive data), a different model (binomial for counts of successes, Poisson for rare event counts), or a **simulation-based method** like the bootstrap or a randomization test. Those simulation methods (developed in the randomization-tests and bootstrap-CI chapters) don't require normality --- they let the data speak for themselves.
:::

::: {.workedexample}
A hospital measures the time patients wait to be seen in the emergency room. The data range from 3 minutes to 6 hours, with most patients waiting under 45 minutes but a long right tail. A researcher wants to run a one-sample $t$-test to see if the mean wait time exceeds 30 minutes. Is a $t$-test appropriate?

---

The data are strongly right-skewed and bounded below by zero, so they are clearly not normal. A one-sample $t$-test *may* still be reasonable if the sample size is large --- the CLT says the sampling distribution of $\bar{x}$ will be approximately normal even when the population is skewed. But with a small sample (say, $n < 30$), the $t$-test could give misleading results. A safer approach is a bootstrap confidence interval for the mean wait time, which doesn't rely on normality of the population.
:::

::: {.statlens}
**See it in action.** Open the [Descriptive Statistics Explorer]({{< var tool-url >}}/explore/descriptive/){target="_blank"} on any bundled dataset --- switch to the histogram view to eyeball symmetry and unimodality, then check the summary statistics for extreme outliers. For a formal look, StatLens's Normal Distribution Explorer's overlay mode fits a normal curve to your data so you can see where the fit is good and where it fails.
:::


## Chapter review {#sec-normal-distribution-review}

### Summary

- Before applying the normal model, **check whether the data are approximately normal**. Two visual tools help: a **histogram with a normal-curve overlay**, and a **QQ plot** (points near the diagonal = approximately normal; systematic curvature = departure from normality).
- The four QQ-plot patterns worth recognizing: **normal** (straight), **right-skewed** (curves up at the right), **left-skewed** (curves down at the left), and **heavy-tailed** (curves away at both ends).
- The normal model fits **symmetric, unimodal, bell-shaped** variables without extreme outliers. Standardized test scores, adult heights, and measurement errors are canonical examples.
- The normal model does **not** fit strongly skewed data (income, waiting times), bounded data (proportions, ages), or count data (better handled by the binomial or Poisson).
- When the normal model doesn't fit, use a **simulation-based** method (bootstrap CI, randomization test) or a **different distribution family** appropriate for the data.

### Key terms

QQ plot (normal probability plot), normality assessment, histogram-with-normal-overlay, symmetric / unimodal / bell-shaped, heavy-tailed distribution, right-skewed / left-skewed, bootstrap alternative to normal-based inference.

## Exercises {.exercises}

Answers to odd-numbered exercises are provided in the [Exercise Solutions appendix](../appendices/exercise-solutions.qmd#sec-solutions-normal-distribution) at the back of the book.

{{< include ../exercises/_normal-distribution-exercises.qmd >}}
