# Tech Tutorial: Simulate {.unnumbered}

::: {.project-meta}
**Tech Tutorial — Simulate.** This unit's tools build a sampling distribution by
**resampling** — shuffling for a hypothesis test, bootstrapping for a confidence
interval. Pick **one** tool and follow its tab.

- **StatLens** — purpose-built for resampling; the natural home for this unit.
- **R** — the `infer` package mirrors the same logic in code.
- **Jamovi** — its standard menus do *not* do randomization/bootstrap (see its tab).

We reuse the class survey from the Explore tutorial: [`class_survey.csv`]({{< var data-url >}}/class_survey.csv){download="class_survey.csv"}.
:::

## What you'll do {#sec-tech-sim-skills}

1. A **randomization test** — is mean `commute_min` different for On- vs. Off-campus students?
2. A **bootstrap confidence interval** — estimate the mean `sleep_hours`.

---

## 1. Randomization test for a difference in means {#sec-tech-sim-rand}

We ask whether on-campus and off-campus students differ in mean one-way commute.
The **null** says housing doesn't matter, so any labeling of students as
on/off-campus is interchangeable. We **shuffle** the housing labels many times to
build the null distribution of the difference in means, then see where the
observed difference falls.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Randomization: Difference in Means tool]({{< var tool-url >}}/simulate/randomization-diff-means/?csv={{< var data-url >}}/class_survey.csv){target="_blank"}.
2.  Set the **response** to `commute_min` and the **grouping** variable to `housing`.
3.  Generate ~1000 shuffles. Read the **p-value** — the share of shuffles at least
    as extreme as the observed difference.

### R

```{r}
#| label: tech-sim-rand
#| echo: true
#| message: false
#| warning: false
#| fig-height: 3
#| fig-cap: "Randomization (null) distribution of the difference in mean commute. The observed difference sits far in the tail."
#| fig-alt: "A bell-shaped null distribution centered at zero with the observed difference marked well into the right tail."

library(readr)
library(dplyr)
library(infer)
library(ggplot2)

survey <- read_csv("../datasets/class_survey.csv")

obs_diff <- survey |>
  specify(commute_min ~ housing) |>
  calculate(stat = "diff in means", order = c("Off-campus", "On-campus"))

null_dist <- survey |>
  specify(commute_min ~ housing) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c("Off-campus", "On-campus"))

null_dist |> get_p_value(obs_stat = obs_diff, direction = "two-sided")

null_dist |>
  visualize() +
  shade_p_value(obs_stat = obs_diff, direction = "two-sided")
```

### Jamovi

<!-- MADDIE (Jamovi owner): esci / BrawStats add-on modules offer *menu-driven*
sampling-variability and null-distribution demos, but NOT a general permutation
p-value on loaded data. If you want a Jamovi-only simulation path here, evaluate
esci and add a click-path. See docs/jamovi-screenshots-todo.md. -->
::: {.callout-note}
## Jamovi: built-in menus don't shuffle

Jamovi's *built-in* menus don't run randomization (permutation) tests — a general
permutation test on your data needs the **StatLens** tab (no install) or the **R**
tab. (Add-on modules such as **esci** give menu-driven *sampling-variability*
demos, but not a permutation p-value.) Jamovi returns in force in the next unit,
*Calculate*, with the formula-based version of this comparison — a two-sample
*t*-test.
:::

:::

---

## 2. Bootstrap confidence interval for a mean {#sec-tech-sim-boot}

To estimate the mean nightly sleep with a margin of error, we **bootstrap**:
resample the data *with replacement* many times, record each resample's mean, and
take the middle 95% of those means as the interval.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Bootstrap: One Mean tool]({{< var tool-url >}}/simulate/bootstrap-mean/?csv={{< var data-url >}}/class_survey.csv){target="_blank"}.
2.  Choose `sleep_hours`. Generate ~1000 bootstrap resamples.
3.  Read the **95% confidence interval** (the 2.5th–97.5th percentiles).

### R

```{r}
#| label: tech-sim-boot
#| echo: true
#| message: false
#| warning: false
#| fig-height: 3
#| fig-cap: "Bootstrap distribution of the mean sleep hours, with the 95% percentile interval shaded."
#| fig-alt: "A roughly normal bootstrap distribution of sample means with the central 95 percent shaded."

boot_dist <- survey |>
  specify(response = sleep_hours) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "mean")

ci <- boot_dist |> get_confidence_interval(level = 0.95, type = "percentile")
ci

boot_dist |>
  visualize() +
  shade_confidence_interval(endpoints = ci)
```

### Jamovi

::: {.callout-note}
## Jamovi: bootstrapping needs a module

A bootstrap CI for a plain mean isn't in Jamovi's built-in menus — use the
**StatLens** or **R** tab. (The **esci** add-on does menu-driven interval
estimation and the "dance of the means"; **GAMLj** does true bootstrap CIs for
regression/ANOVA models.) In the *Calculate* unit, Jamovi produces the
formula-based one-sample *t* interval for this same mean.
:::

:::

---

## Check yourself {#sec-tech-sim-check}

- [ ] run a randomization test for a difference and read its p-value;
- [ ] build and read a bootstrap confidence interval;
- [ ] explain what is being shuffled (test) vs. resampled with replacement (interval).

::: {.callout-tip collapse="true"}
## Going further in R (optional)

OpenIntro *IMS* interactive tutorials for this unit (same `infer` style as above):

- [Sampling variability](https://openintro.shinyapps.io/ims-04-foundations-01/){target="_blank"}
- [Randomization test](https://openintro.shinyapps.io/ims-04-foundations-02/){target="_blank"}
- [Bootstrapping](https://openintro.shinyapps.io/ims-05-infer-05/){target="_blank"}
:::
