# Tech Tutorial: Compute {.unnumbered}

::: {.project-meta}
**Tech Tutorial — Compute.** This unit uses **formulas and theoretical
distributions** (the *t* and the normal) instead of resampling. Pick **one** tool
and follow its tab — all three handle this unit well.

We reuse [`class_survey.csv`]({{< var data-url >}}/class_survey.csv){download="class_survey.csv"}.
:::

## What you'll do {#sec-tech-calc-skills}

1. A **two-sample *t*-test** — do On- and Off-campus students differ in mean `sleep_hours`?
2. A **confidence interval for one proportion** — what fraction of students live on campus?

---

## 1. Two-sample *t*-test for a difference in means {#sec-tech-calc-t}

This is the formula-based counterpart to the randomization test from the Simulate
unit. It assumes roughly normal data (or a large enough sample) and reports a
*t* statistic, degrees of freedom, and a p-value.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Two-Means Inference tool]({{< var tool-url >}}/inference/two-means/?csv={{< var data-url >}}/class_survey.csv&group=housing&response=sleep_hours){target="_blank"}.
2.  Confirm the **grouping** variable is `housing` and the **response** is `sleep_hours`.
3.  Read the **t statistic, df, p-value**, and the confidence interval for the difference.

### Jamovi

1.  Open `class_survey.csv`.
2.  **T-Tests → Independent Samples T-Test.** Dependent: `sleep_hours`; Grouping: `housing`.
3.  Tick **Welch's** (unequal variances), **Mean difference**, and **Confidence interval**.

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-calc-t
#| echo: true
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(infer)

survey <- read_csv("../datasets/class_survey.csv")

# Two-sample t-test (Welch); order sets the sign of the difference
survey |>
  t_test(sleep_hours ~ housing,
         order = c("On-campus", "Off-campus"))
```

:::

---

## 2. Confidence interval for one proportion {#sec-tech-calc-prop}

Here the variable is categorical: each student lives on campus or not. We estimate
the population proportion living on campus with a normal-based (Wald/score)
confidence interval.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [One-Proportion Inference tool]({{< var tool-url >}}/inference/one-prop/?csv={{< var data-url >}}/class_survey.csv){target="_blank"}.
2.  Choose `housing` and set the **success** category to `On-campus`.
3.  Read the sample proportion and its **95% confidence interval**.

### Jamovi

1.  **Frequencies → 2 Outcomes (Binomial test).** Variable: `housing`.
2.  Set the **test value** and read the **proportion** and its **confidence interval**
    (Jamovi reports the CI for the level you select).

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-calc-prop
#| echo: true
#| message: false
#| warning: false

# Count on-campus students and form a one-proportion z interval
on <- sum(survey$housing == "On-campus")
n  <- nrow(survey)

prop.test(on, n)$conf.int        # 95% CI for the proportion on campus
round(on / n, 3)                 # sample proportion
```

:::

---

## Check yourself {#sec-tech-calc-check}

- [ ] run a two-sample *t*-test and report t, df, p-value, and the CI for the difference;
- [ ] build a confidence interval for a single proportion;
- [ ] state, for each, the condition that makes the theoretical model appropriate.

::: {.callout-tip collapse="true"}
## Going further in R (optional)

OpenIntro *IMS* interactive tutorials for this unit (same `infer` style):

- [Inference for a single proportion](https://openintro.shinyapps.io/ims-05-infer-01/){target="_blank"}
- [Inference for comparing two proportions](https://openintro.shinyapps.io/ims-05-infer-02/){target="_blank"}
- [Inference for difference in two means](https://openintro.shinyapps.io/ims-05-infer-07/){target="_blank"}
- [Introducing the t-distribution](https://openintro.shinyapps.io/ims-05-infer-06/){target="_blank"}
:::
