# Tech Tutorial: Apply {.unnumbered}

::: {.project-meta}
**Tech Tutorial — Apply.** This unit extends inference to more than two groups and
to relationships between variables: **chi-square**, **ANOVA**, and **regression**.
Pick **one** tool and follow its tab.

We reuse [`class_survey.csv`]({{< var data-url >}}/class_survey.csv){download="class_survey.csv"}.
:::

## What you'll do {#sec-tech-apply-skills}

1. A **chi-square test of independence** — are `year` and `housing` associated?
2. A **one-way ANOVA** — does mean `sleep_hours` differ across class `year`?
3. A **simple linear regression** — does `study_hours` predict `sleep_hours`?

---

## 1. Chi-square test of independence {#sec-tech-apply-chisq}

Both variables are categorical, so we cross-tabulate `year` × `housing` and test
whether the two are independent.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Chi-Square Independence tool]({{< var tool-url >}}/inference/chisq/?csv={{< var data-url >}}/class_survey.csv&group=year&response=housing){target="_blank"}.
2.  Confirm the two variables are `year` and `housing`.
3.  Read the **contingency table, expected counts, χ² statistic, df, and p-value**.

### Jamovi

1.  Open `class_survey.csv`.
2.  **Frequencies → Independent Samples (χ² test of association).** Rows: `year`; Columns: `housing`.
3.  Under **Statistics**, read **χ²** and **p**; under **Cells**, request **Expected counts**.

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-apply-chisq
#| echo: true
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(infer)

survey <- read_csv("../datasets/class_survey.csv")

survey |> chisq_test(housing ~ year)
```

:::

::: {.callout-warning}
## Check the conditions

With only 50 students split across a 4×2 table, some **expected counts fall below
5** — the chi-square approximation is shaky here. In practice you'd report that
caveat (and could fall back to the *randomization* chi-square from the Simulate
unit). For this tutorial the point is the mechanics.
:::

---

## 2. One-way ANOVA {#sec-tech-apply-anova}

ANOVA compares the mean of a numerical variable (`sleep_hours`) across 3+ groups
(the four class years) with a single *F*-test.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [One-Way ANOVA tool]({{< var tool-url >}}/inference/anova/?csv={{< var data-url >}}/class_survey.csv&group=year&response=sleep_hours){target="_blank"}.
2.  Confirm **response** `sleep_hours`, **grouping** `year`.
3.  Read the group means, the **F statistic**, df, and **p-value**.

### Jamovi

1.  **ANOVA → One-Way ANOVA.** Dependent: `sleep_hours`; Grouping: `year`.
2.  Read the **F**, **df**, and **p**. (Use **Descriptives plots** to see the group means.)

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-apply-anova
#| echo: true
#| message: false
#| warning: false

aov(sleep_hours ~ year, data = survey) |> summary()
```

:::

---

## 3. Simple linear regression {#sec-tech-apply-reg}

Both variables are numerical. We fit a line predicting `sleep_hours` from
`study_hours` and test whether the slope differs from zero. (In the Explore unit
the scatter of these two looked flat — regression now tests that formally.)

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Regression Slope tool]({{< var tool-url >}}/inference/slope/?csv={{< var data-url >}}/class_survey.csv){target="_blank"}.
2.  Set **x** to `study_hours` and **y** to `sleep_hours`.
3.  Read the fitted **slope**, its **p-value**, and **R²**.

### Jamovi

1.  **Regression → Linear Regression.** Dependent: `sleep_hours`; Covariate: `study_hours`.
2.  Read the **slope** (the `study_hours` coefficient), its **p**, and the **R²** (Model Fit).

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-apply-reg
#| echo: true
#| message: false
#| warning: false

lm(sleep_hours ~ study_hours, data = survey) |> summary()
```

:::

---

## Check yourself {#sec-tech-apply-check}

- [ ] run a chi-square test of independence and check its expected counts;
- [ ] run a one-way ANOVA and report F, df, and p;
- [ ] fit a simple regression and report the slope, its p-value, and R²;
- [ ] match each procedure to the data shape that calls for it.

::: {.callout-tip collapse="true"}
## Going further in R (optional)

OpenIntro *IMS* interactive tutorials for this unit:

- [Chi-squared test of independence](https://openintro.shinyapps.io/ims-05-infer-03/){target="_blank"}
- [Comparing many means (ANOVA)](https://openintro.shinyapps.io/ims-05-infer-08/){target="_blank"}
- [Simple linear regression](https://openintro.shinyapps.io/ims-03-model-03/){target="_blank"}
- [Inference for regression (t-test for slope)](https://openintro.shinyapps.io/ims-06-model-infer-03/){target="_blank"}
:::
