# Tech Tutorial: Foundations {.unnumbered}

::: {.project-meta}
**Tech Tutorial — Foundations.** This unit works with **probability distributions**
directly — no data file. You give a distribution its parameters and read off
probabilities and percentiles. Pick **one** tool and follow its tab.

To keep things concrete we imagine the *population* behind the class survey:
study hours across all intro-stats students are roughly **Normal, mean 13, SD 4**,
and a fraction **0.44** of all students live on campus.
:::

## What you'll do {#sec-tech-found-skills}

1. **Normal** probabilities and percentiles.
2. **Binomial** probabilities.

---

## 1. Normal distribution {#sec-tech-found-normal}

Take study hours to be $X \sim N(\mu = 13,\ \sigma = 4)$. Find:

- **(a)** the probability a student studies **more than 20 hours**, $P(X > 20)$;
- **(b)** the **90th percentile** of study hours.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Normal Distribution calculator]({{< var tool-url >}}/distribution/normal/?mu=13&sigma=4){target="_blank"} (mean and SD are pre-filled).
2.  For **(a)**, set the boundary to **20** and select the **upper tail** — read the shaded probability.
3.  For **(b)**, switch to **inverse / find-x** mode and enter **0.90** to get the percentile.

### Jamovi

1.  Install the **distrACTION** module (Jamovi library), then **distrACTION → Normal Distribution**.
2.  Enter **Mean = 13**, **SD = 4**.
3.  Use **Compute probability** with $x_1 = 20$ and the upper tail for **(a)**;
    use **Compute quantile** with $p = 0.90$ for **(b)**.

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-found-normal
#| echo: true

# (a) P(X > 20) for X ~ N(13, 4)
pnorm(20, mean = 13, sd = 4, lower.tail = FALSE)

# (b) 90th percentile
qnorm(0.90, mean = 13, sd = 4)
```

:::

---

## 2. Binomial distribution {#sec-tech-found-binom}

Suppose **44%** of all students live on campus. In a random sample of $n = 12$
students, let $X$ be the number living on campus, so $X \sim \text{Binomial}(n = 12,\ p = 0.44)$. Find:

- **(a)** $P(X = 6)$ — exactly six;
- **(b)** $P(X \ge 6)$ — six or more.

::: {.panel-tabset group="tool"}

### StatLens

1.  Open the [Binomial Distribution calculator]({{< var tool-url >}}/distribution/binomial/){target="_blank"}.
2.  Set **n = 12** and **p = 0.44**.
3.  For **(a)** read the bar at **k = 6**; for **(b)** select the **≥ 6** (upper) region.

### Jamovi

1.  **distrACTION → Binomial Distribution** (same module as above).
2.  Enter **n = 12**, **p = 0.44**.
3.  **Compute probability** with $x = 6$ for **(a)**, and the **$P(X \ge x)$** option for **(b)**.

*(Jamovi screenshots to be added.)*

### R

```{r}
#| label: tech-found-binom
#| echo: true

# (a) P(X = 6) for X ~ Binomial(12, 0.44)
dbinom(6, size = 12, prob = 0.44)

# (b) P(X >= 6) = 1 - P(X <= 5)
pbinom(5, size = 12, prob = 0.44, lower.tail = FALSE)
```

:::

---

## Check yourself {#sec-tech-found-check}

- [ ] find a tail probability and a percentile for a normal distribution;
- [ ] find an exact and a cumulative probability for a binomial distribution;
- [ ] choose the right distribution: normal for a measured amount, binomial for a count of successes in $n$ trials.

::: {.callout-tip collapse="true"}
## Going further in R (optional)

The OpenIntro *IMS* and *OpenIntro Statistics* materials cover the probability
foundations; the R functions above (`pnorm`, `qnorm`, `dbinom`, `pbinom`) are the
core toolkit. See the [normal](https://openintro.shinyapps.io/ims-05-infer-06/){target="_blank"}
tutorial for how the normal model underlies later inference.
:::
