---
title: "Statistical Tables"
---

```{r}
#| include: false
#| cache: false
library(dplyr)
library(kableExtra)
library(tibble)
knitr::opts_chunk$set(echo = FALSE)
```

Standard reference tables for the four distributions used in this textbook: **standard normal** ($z$), **Student's $t$**, **chi-square** ($\chi^2$), and **$F$**. Use them for exams or offline work; for everyday computation, use StatLens's [distribution calculators](https://learnlens.org/statlens/distribution/), R, or Jamovi.

## Common critical values --- cheat sheet {#sec-crit-values-cheat}

The values below appear most often when constructing confidence intervals or two-sided tests. Read across each row: for a 95% CI or $\alpha = 0.05$ two-sided test, use 1.96 for $z^{*}$; use the $t^{*}$ row for the appropriate $df$. The last row shows the large-sample limit $t^{*} \to z^{*}$.

```{r}
#| label: tbl-crit-cheat
#| tbl-cap: "Common two-sided critical values (t-star, and z-star in the last row)."
levels <- c("90%", "95%", "98%", "99%")
alpha  <- c(0.10, 0.05, 0.02, 0.01)
z_star <- qnorm(1 - alpha / 2)

t_star <- function(df) qt(1 - alpha / 2, df)
dfs <- c(5, 10, 20, 30, 50, 100, 1000)
t_rows <- sapply(dfs, t_star)
t_df <- as.data.frame(t(t_rows))
colnames(t_df) <- levels
t_df <- tibble(`df` = as.character(dfs)) |> bind_cols(t_df)

# z-star row: plain-text label (kable captions/cells don't render inline math)
z_row <- tibble(`df` = "∞ (z*)", `90%` = z_star[1], `95%` = z_star[2],
                `98%` = z_star[3], `99%` = z_star[4])

crit <- t_df |>
  mutate(across(-df, ~ formatC(.x, digits = 3, format = "f"))) |>
  bind_rows(z_row |> mutate(across(-df, ~ formatC(.x, digits = 3, format = "f"))))

kbl(crit, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Standard normal (z) {#sec-z-table}

Entries are cumulative left-tail probabilities $P(Z \le z)$ for the standard normal distribution. Read the leading digit and first decimal from the row label, then the second decimal from the column. For right-tail probabilities use $1 - P(Z \le z)$; for two-tailed use symmetry.

```{r}
#| label: tbl-z-table
#| tbl-cap: "Cumulative standard normal probabilities."
z_row  <- seq(-3.4, 3.4, by = 0.1)
z_col  <- seq(0.00, 0.09, by = 0.01)
zvals  <- outer(z_row, z_col, FUN = "+")
probs  <- pnorm(zvals)

tbl <- as.data.frame(round(probs, 4))
colnames(tbl) <- sprintf("0.%02d", 0:9)
rownames(tbl) <- sprintf("%+.1f", z_row)

kbl(tbl, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 10, full_width = FALSE) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```

## Student's t {#sec-t-table}

Entries are critical values $t^{*}$ such that $P(T > t^{*}) = \alpha_{\text{right-tail}}$ for a $t$-distribution with the row's degrees of freedom. For a two-sided CI at confidence level $C$, use the column headed $\alpha_{\text{two-tail}} = 1 - C$.

**Row**: degrees of freedom. **Columns**: one-tail area $\alpha$ to the right of $t^{*}$.

```{r}
#| label: tbl-t-table
#| tbl-cap: "Upper-tail critical values of the t-distribution."
one_tail <- c(0.10, 0.05, 0.025, 0.01, 0.005, 0.001)
dfs <- c(1:30, 40, 50, 60, 80, 100, 200, 500, 1000)

crit <- sapply(one_tail, function(a) qt(1 - a, dfs))
tbl <- as.data.frame(round(crit, 3))
colnames(tbl) <- sprintf("%.3f", one_tail)
tbl <- cbind(df = dfs, tbl)

kbl(tbl, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 10, full_width = FALSE)
```

::: {.callout-tip}
## Reading the t-table for a confidence interval

For a two-sided $C\%$ confidence interval, use one-tail area $\alpha = (1 - C/100) / 2$. So a 95% CI uses the $\alpha = 0.025$ column; a 99% CI uses the $\alpha = 0.005$ column.
:::

## Chi-square ($\chi^2$) {#sec-chisq-table}

Entries are critical values $\chi^{2\,*}$ such that $P(X^{2} > \chi^{2\,*}) = \alpha$ for a $\chi^2$-distribution with the row's degrees of freedom.

**Row**: degrees of freedom. **Columns**: right-tail area $\alpha$ (i.e., $P(X^{2} > \chi^{2\,*}) = \alpha$).

```{r}
#| label: tbl-chisq-table
#| tbl-cap: "Upper-tail critical values of the chi-square distribution."
alphas <- c(0.995, 0.99, 0.975, 0.95, 0.90, 0.10, 0.05, 0.025, 0.01, 0.005)
dfs <- c(1:30, 40, 50, 60, 80, 100)

crit <- sapply(alphas, function(a) qchisq(1 - a, dfs))
tbl <- as.data.frame(round(crit, 3))
colnames(tbl) <- sprintf("%.3f", alphas)
tbl <- cbind(df = dfs, tbl)

kbl(tbl, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 10, full_width = FALSE)
```

## $F$-distribution {#sec-f-table}

Entries are critical values $F^{*}$ such that $P(F > F^{*}) = \alpha$ for an $F$-distribution with numerator degrees of freedom $df_1$ (columns) and denominator degrees of freedom $df_2$ (rows). Separate tables for $\alpha = 0.05$ and $\alpha = 0.01$.

### $\alpha = 0.05$

**Rows**: denominator df ($df_2$). **Columns**: numerator df ($df_1$).

```{r}
#| label: tbl-f-05
#| tbl-cap: "F-star with right-tail area 0.05."
df1s <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 12, 15, 20, 30, 60, 120)
df2s <- c(1:20, 25, 30, 40, 60, 120, 1000)
crit <- outer(df2s, df1s, function(d2, d1) qf(0.95, d1, d2))
tbl <- as.data.frame(round(crit, 2))
colnames(tbl) <- df1s
tbl <- cbind(df2 = df2s, tbl)

kbl(tbl, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 9, full_width = FALSE)
```

### $\alpha = 0.01$

**Rows**: denominator df ($df_2$). **Columns**: numerator df ($df_1$).

```{r}
#| label: tbl-f-01
#| tbl-cap: "F-star with right-tail area 0.01."
crit <- outer(df2s, df1s, function(d2, d1) qf(0.99, d1, d2))
tbl <- as.data.frame(round(crit, 2))
colnames(tbl) <- df1s
tbl <- cbind(df2 = df2s, tbl)

kbl(tbl, align = "c", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 9, full_width = FALSE)
```

## Alternatives

- **StatLens distribution calculators** at [learnlens.org/statlens/distribution/](https://learnlens.org/statlens/distribution/) --- interactive versions of the same distributions with visual tail-shading.
- **R**: `pnorm`, `qnorm`, `pt`, `qt`, `pchisq`, `qchisq`, `pf`, `qf` (the `p*` functions give left-tail probabilities; the `q*` functions give quantiles/critical values).
- **Jamovi**: use the `descriptives` panel with the `p-value` and `quantile` options, or the distributions module.
